---
title: "setup"
author: "Phillip Hungerford"
date: "Monday 18th November 2024"
output: html_document
---

```{r setup, include=FALSE}
options(repos = c(CRAN = 'https://cran.rstudio.com'))
#install.packages('knitr')
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

Keras is an API that runs tensorflow, JAX and pytorch as backends. 

To quickly get up and running, it is recommended to use the included Dockerfile, this sets up all dependencies for keras and gets it working 'out of the box'. 

For installation without using the Dockerfile refer to the [Keras3](https://keras3.posit.co/articles/getting_started.html#installation).

# Test

Run the small test model below to see if your keras has been setup correctly. 

```{r}

library(keras3)

# Generate dummy data
x <- matrix(runif(1000), nrow = 100, ncol = 10)
y <- sample(0:1, 100, replace = TRUE)

# Define a model
model <- keras_model_sequential() %>%
  layer_dense(units = 32, activation = 'relu', input_shape = c(10)) %>%
  layer_dense(units = 1, activation = 'sigmoid')

# Compile the model
model %>% compile(
  optimizer = 'adam',
  loss = 'binary_crossentropy',
  metrics = c('accuracy')
)

# Fit the model
model %>% fit(x, y, epochs = 10, batch_size = 32)
```

